Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-02-27 write-ups/FSE/paper.tex
Fixed some spelling issues.

For example, Nvidia's latest architecture (Pascal) is documented to
support instruction-level preemption~\cite{PascalWhitepaper}. Despite
this, Nvidia's most recent compute capablity does not guarantee
forward progress for CUDA programmers (evidence that vendors will not
yet commit to providing such guarantees).  Indeed, on a GTX Titan X
(Pascal) we still observe starvation: we found that a global barrier
can be executed successfully with 56 workgroups, but deadlocks with 57
workgroups.  This is an example of a vendor supporting preemption, yet
not guaranteeing fair scheduling.

@@

The semantics of $\resizingglobalbarrier$ can be modelled via calling $\offerfork$ and $\offerkill$
surrounded, and seperated by calls to a $\globalbarrier$.
%
%\lstset{basicstyle=\small\tt,numbers=none}
%\begin{lstlisting}
%  $\globalbarrier()$;  $\offerfork$();
%  $\globalbarrier()$; $\offerkill()$;
%  $\globalbarrier()$;
%\end{lstlisting}
%\lstset{basicstyle=\scriptsize\tt,numbers=left}
%
The enclosing $\globalbarrier$ calls ensure that the change in number
of active workgroups from $M$ to $M'$ occurs entirely within the
resizing barrier, so that from a programmer's perspective, $M$ changes atomically.  The middle $\globalbarrier$ ensures that forking occurs before killing, so that workgroups $[0, \textrm{min}(M, M') - 1]$ are left intact.

@@

The middle and right graphs show results for graph algorithms.  These
algorithms use barriers, and we experimented with the regular and
query barrier implementations described in
\mysec\ref{sec:resizingbarrier}.  The execution times for the
non-cooperative task are averaged across all runs, including with both
types of barrier.  We show separately the average gather time
associated with each type of barrier.  The graphs show a similar trend
to the left-most graph: as the number of non-cooperative workgroups
grows, the execution time decreases and the gather time
increases.%\footnote{The right figure has only 3 points per line:
  %recall that for bfs there are no runs with ``all-but-one''
  %non-cooperative workgroups.}
%
the gather time is higher on the right figure: the G3 circuit input
graph is rather wide than deep, so the graph algorithm reaches
resizing barriers less often than for the usa road input of the middle
figure for instance. The scheduler thus has fewer opportunities to
collect workgroups and gather time increases. Nonetheless, scheduling
responsiveness can benefit from the query barrier: when used, this
barrier lets the scheduler collect all needed workgroups as soon as
they hit a resizing barrier. As we can see, the gather time of the
query barrier is almost stable w.r.t.\ the number of workgroups that
needs to be collected.

@@

\myparagraph{GPU multitasking and scheduling}
%
%\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
%
Hardware support for preemption has been proposed for \nvidia GPUs, as
well as \emph{SM-draining} whereby workgroups occupying a symmetric
multiprocessor (SM; a compute unit using our terminology) are allowed
to complete until the SM becomes free for other
tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited
the presence of blocking constructs, since it may not be possible to
drain a blocked workgroup.
%
A follow-up work adds the notion of SM \emph{flushing}, where a
workgroup can be re-scheduled from scratch if it has not yet committed
side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have
been evaluated using simulators, over sets of regular GPU kernels.
Very recent \nvidia GPUs (i.e. the Pascal architecture) support
preemption, though, as discussted in \mysec{\ref{sec:intro}} and
\mysec{\ref{sec:nvidiacomparison}}, it is not clear whether they guarantee
fairness or allow tasks to share GPU resources at the workgroup
level~\cite{PascalWhitepaper}.

>>>>

For example, Nvidia's latest architecture (Pascal) is documented to
support instruction-level preemption~\cite{PascalWhitepaper}. Despite
this, Nvidia's most recent compute capability does not guarantee
forward progress for CUDA programmers (evidence that vendors will not
yet commit to providing such guarantees).  Indeed, on a GTX Titan X
(Pascal) we still observe starvation: we found that a global barrier
can be executed successfully with 56 workgroups, but deadlocks with 57
workgroups.  This is an example of a vendor supporting preemption, yet
not guaranteeing fair scheduling.

@@

The semantics of $\resizingglobalbarrier$ can be modelled via calling $\offerfork$ and $\offerkill$
surrounded, and separated by calls to a $\globalbarrier$.
%
%\lstset{basicstyle=\small\tt,numbers=none}
%\begin{lstlisting}
%  $\globalbarrier()$;  $\offerfork$();
%  $\globalbarrier()$; $\offerkill()$;
%  $\globalbarrier()$;
%\end{lstlisting}
%\lstset{basicstyle=\scriptsize\tt,numbers=left}
%
The enclosing $\globalbarrier$ calls ensure that the change in number
of active workgroups from $M$ to $M'$ occurs entirely within the
resizing barrier, so that from a programmer's perspective, $M$ changes atomically.  The middle $\globalbarrier$ ensures that forking occurs before killing, so that workgroups $[0, \textrm{min}(M, M') - 1]$ are left intact.

@@

The middle and right graphs show results for graph algorithms.  These
algorithms use barriers, and we experimented with the regular and
query barrier implementations described in
\mysec\ref{sec:resizingbarrier}.  The execution times for the
non-cooperative task are averaged across all runs, including with both
types of barrier.  We show separately the average gather time
associated with each type of barrier.  The graphs show a similar trend
to the left-most graph: as the number of non-cooperative workgroups
grows, the execution time decreases and the gather time
increases.%\footnote{The right figure has only 3 points per line:
  %recall that for bfs there are no runs with ``all-but-one''
  %non-cooperative workgroups.}
%
the gather time is higher on the right figure: the G3 circuit input
graph is rather wide than deep, so the graph algorithm reaches
resizing barriers less often than for the USA road input of the middle
figure for instance. The scheduler thus has fewer opportunities to
collect workgroups and gather time increases. Nonetheless, scheduling
responsiveness can benefit from the query barrier: when used, this
barrier lets the scheduler collect all needed workgroups as soon as
they hit a resizing barrier. As we can see, the gather time of the
query barrier is almost stable w.r.t.\ the number of workgroups that
needs to be collected.

@@

\myparagraph{GPU multitasking and scheduling}
%
%\ADComment{Tyler: please update or delete based on your recent readings and refer to Section~\ref{sec:nvidiacomparison}.}
%
Hardware support for preemption has been proposed for \nvidia GPUs, as
well as \emph{SM-draining} whereby workgroups occupying a symmetric
multiprocessor (SM; a compute unit using our terminology) are allowed
to complete until the SM becomes free for other
tasks~\cite{DBLP:conf/isca/TanasicGCRNV14}.  SM draining is limited
the presence of blocking constructs, since it may not be possible to
drain a blocked workgroup.
%
A follow-up work adds the notion of SM \emph{flushing}, where a
workgroup can be re-scheduled from scratch if it has not yet committed
side-effects~\cite{DBLP:conf/asplos/ParkPM15}.  Both approaches have
been evaluated using simulators, over sets of regular GPU kernels.
Very recent \nvidia GPUs (i.e. the Pascal architecture) support
preemption, though, as discussed in \mysec{\ref{sec:intro}} and
\mysec{\ref{sec:nvidiacomparison}}, it is not clear whether they guarantee
fairness or allow tasks to share GPU resources at the workgroup
level~\cite{PascalWhitepaper}.
